---
title: "Experiment_1_2"
format: html
editor: visual
---

## Analysis of inoculum on fungal growth

This data set compares fungal growth speed and bacterial log-change for a set of fungal species able to grow on human faeces.

## Growth speed analysis for inoculum

Table sorting the growth speed according to substrate and inouclum mass.

Compare for single species present data for single species as a table and create a box plot with over view of all species (\~5) for all substrates (aprox. 6)

```{r}

###Prepare tables jpining inoculum_1_2 to exp_1_2

### Normalize growth speed because substrate is never exactly the same 

### Group by species compare within species if there is a difference between the used substrates if there are differences compare what is more relevant substrate used or substrate mass

### Represent data as a boxplot (species wise fititng 4 plots on togetehr )
### showing fungal species as the title
### inoculum on the x-axis as a subcategorie, 
### growth speed on the y-axis as a boxplot
### Mark significant differences with (a,b,c,d)
```

## Bacterial analysis

Compare if the inoculum has an influence on growth speed or bacterial log-change.

```{r}

### Prepare data tables joining experimen_set_up, bacterial analysis, and inoculum used

### Calculate log-change for each repllicate

### Group by species

### Test for normality

### (Possible other tests)

### Compare e_coli and enterococcus log-change (ANOVA oder ANCOVA)


### Prepare plots for each species an individual 






```

## 
